How to make an abridged life table and calculate life expectancy

### 2016-09-19 working_teaching_examples/how_to_abridged_life_table.Rmd
change life expectancy calc 2

As a practical alternative, a life table can be made using current mortality rates for a population. This is called a current life table [@selvin_text].

@@

6. Total time lived ($T_x$): This is the accumulation of person-time-at-risk at age x and beyond -- the total number of person years lived by those alive at age $x$.

@@

##Assumptions of the life table
1. The number of births occurring each year is constant.
2. Deaths are uniformly distributed during the interval after 1 years old.
3. There is no population growth.
Together, these assumptions imply that the population must be stationary.
3.

@@

@@

```{r function_lifeexpectancy}
#dataset needs to be in order of age
life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA, calendar.years = NA){

@@

  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) #probability of dying
  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group
  data["p_x"] <- 1 - data["q_x"] #probability of survival

@@

  #now to calculate the total person time lived at or after age x, T_x:
  data["cumsum"] <- cumsum(data["L_x"])
  data["l.cumsum"] <- c(0, data[1:(dim(data)[1] - 1), "cumsum"])
  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
  data["e_x"] <- data["T_x"]/data["l_x"]

@@

```{r lifetable_cali_abridged}
cali.abridged <- data.frame("ages" = c("0-1y", "1-4y", "5-9y", "10-14y", "15+y"), "nx" = c(1, 4, 5, 5, 5),
                         "obs.deaths" = c(1635, 168, 251, 371, 23300), #last number is fictitious
                         "obs.popn" = c(123342, 435167, 578964, 649693, 143600))

life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "obs.deaths", population.counts = "obs.popn")
```

@@

>>>>

As a practical alternative, a life table can be made using current mortality rates for a population. This is called a current life table [@selvin_text] or a period life table [@preston_demography_text].

@@

6. Cumulative time lived ($T_x$): This is the accumulation of person-time-at-risk at age x and beyond -- the total number of person years lived by those alive at age $x$.

@@

##Assumptions of the life table

1. The number of births occurring each year is constant.
2. Deaths are uniformly distributed during the interval after 1 years old.
3. There is no population growth.

Together, these assumptions imply that the population must be stationary.

@@

**Derivation of the formula for $q_x$ when the life table is abridged**

This derivation was modified from the Preston demography text [@preston_demography_text]. Note that in the Preston book $a_x$ represents *person-years* not the proportion of the year, which is why the formulas look different (because we have to write $n_x\bar{a_x}$ in place of Preston's $a_x$).

We start with the simple formula for q_x. As before:
$q_x=d_x/l_x$.

Now, $L_x = n_x*(l_x - d_x) + n_x\bar{a_x}d_x$
$L_x = n_xl_x - n_xd_x + n_x\bar{a_x}d_x$
$l_x = \frac{L_x + n_xd_x - n_x\bar{a_x}d_x}{n_x}$
$l_x = \frac{L_x + d_xn_x(1 - \bar{a_x}){n_x}$

We can then substitute this expression for $l_x$ into the equations for $q_x$:

$q_x=\frac{d_xn_x}{L_x + d_xn_x(1 - \bar{a_x}}$, and dividing numerator and denominator by L_x, we get
$q_x=\frac{R_xn_x}{1 + R_xn_x(1 - \bar{a_x}}$.

@@

```{r function_lifeexpectancy}
#dataset needs to be in order of age.
life.table <- function(data, age.groups, num.ages.in.group, death.counts, population.counts, ave.prop.lived = NA){

@@

  data["q_x"] <- data[num.ages.in.group]*data["R_x"]/(1 + (1 - data["a_x"])*data[num.ages.in.group]*data["R_x"]) 
  #probability of dying

  data[dim(data)[1], "q_x"] <- 1 #100% die in last age group

  data["p_x"] <- 1 - data["q_x"] #probability of survival

@@

  #now to calculate the total person time lived at or after age x, T_x:
  data["cumsum"] <- cumsum(data["L_x"])
  data["l.cumsum"] <- c(0, unlist(data[1:(dim(data)[1] - 1), "cumsum"]))
  data["T_x"] <- sum(data["L_x"]) - data["l.cumsum"]
  data["e_x"] <- data["T_x"]/data["l_x"]

@@

@@

```{r lifetable_cali_abridged}
cali.complete$group <- cut(cali.complete$ages, breaks=c(0, 1, 5, 10, 15, 20), 
                           labels = c("<1 years old", "1-4 years old", "5-9 years old", 
                                      "10-14 years old", "15 or older"), include.lowest = T, right = F)

table(cali.complete$group, cali.complete$ages)

library(dplyr)
cali.abridged <- cali.complete %>% group_by(group) %>% 
                                   summarise(nx = sum(nx), ages = first(group), popn = sum(obs.popn), deaths = sum(obs.deaths),
                                             avelived = first(avelived))

life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn")

life.table(data = cali.abridged, age.groups = "ages", num.ages.in.group = "nx", 
           death.counts = "deaths", population.counts = "popn", ave.prop.lived = "avelived")
```

```{r}
load("Data/alabama_only.Rdata")

#example data
alabama.by.strata <- 
  dat.clean.alabama %>% group_by(Year2, RaceSex, Age2) %>%
  arrange(RaceSex, Year2, Age2) %>%
  mutate(nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years")) %>%
  summarise(total_deaths5 = sum(Count_md5),
            Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx))

first.strata <- alabama.by.strata[alabama.by.strata$StateYearRaceSex == "Alabama.1969.White.Male", ]

system.time(first.strata.life.table <- life.table(data = first.strata, age.groups = "Age2", num.ages.in.group = "nx", death.counts = "total_deaths5", population.counts = "Population"))

View(first.strata.life.table)

system.time(
for(stratum.id in levels(droplevels(alabama.by.strata$StateYearRaceSex))) {
  assign(paste0("lt.", as.character(stratum.id)), 
         life.table(data = subset(alabama.by.strata, StateYearRaceSex == stratum.id), 
                                  age.groups = "Age2", num.ages.in.group = "nx", 
                                  death.counts = "total_deaths5", population.counts = "Population"))
  print(stratum.id)
}
)
```

Stopped here. For next time: 
need to change T_x to T_x+n in this formula and add the last term in the empty brackets after leading the number alive variable
see the auger paper pg 576 for the formula to finish programming.
data1 <- Quebec
data2 <- Canada

```{r function_le_age_decomp}

le_age_decomp <- function(data1, data2, num.alive, interval.years.lived, accumulated.years.lived) {

  C_x <- 
  (data1[num.alive]/data1[1, num.alive])*((data2[interval.years.lived]/data2[num.alive]) - (data1[interval.years.lived]/data1[num.alive])) +
    ((data2[accumulated.years.lived]/data2[1, num.alive])*((data1[num.alive]/data2[num.alive]) - ))
}

```
